---
title: "R Notebook"
date: "11/12/2019"
output:
  github_document:
    html_preview: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load data, calculate mean diastolic and systolic blood pressures, and removing individual blood pressure readings and other non-essential variables
```{r load-data, eval=TRUE}

bp = read.csv("NHANES Filtered Workbook.csv")
bp$BPAvgSys = rowMeans(bp[c("BPXSY1", "BPXSY2", "BPXSY3")], na.rm=TRUE)
bp$BPAvgDias = rowMeans(bp[c("BPXDI1", "BPXDI2", "BPXDI3")], na.rm=TRUE)
library(dplyr)
bp = select(bp,-c(SEQN, BPXSY1, BPXSY2, BPXSY3,BPXDI1,BPXDI2, BPXDI3, LBXIN, LBDINSI, LBDLDL, DIQ050, URXPREG,IND235))
```

### Ommiting any nulls in the dataset and replacing values in DBD895 (# of meals not prepared at home) with 22 where it shows value "5,555". This is an obvious error in the data, so we replaced with average for the column.
```{r}
bp = na.omit(bp)
bp$DBD895 = replace(bp$DBD895, bp$DBD895==5555, 22)
```



### Filtering down the dataset to remove dysfunctional data
```{r}
bp = bp %>% filter(RIDEXPRG != 3, INDHHIN2 %in% c(1:10,14,15), DRQSPREP != 9, DRQSDIET != 9, !ALQ130 %in% c(777,999), !DIQ010 %in% c(7,9), !DIQ160 %in% c(7,9), !DBD895 %in% c(7777,9999), !DPQ030 %in% c(7,9), !MCQ080 %in% c(7,9), !MCQ160B %in% c(7,9), !MCQ160C %in% c(7,9), !MCQ160F %in% c(7,9), !PAD680 %in% c(7777,9999), !SMQ020 %in% c(7,9))
```


### Renaming columns
```{r}
colnames(bp)[1:34] = c("Sex","Age","Race","Pregnant", "InfantsInHousehold", "ChildrenInHousehold", "SeniorsInHousehold", "Income", "SaltUsed", "SpecialDiet", "Calcium","Magnesium","Sodium","Potassium","Alcohol", "Weight","BMI", "Waist", "LabCalcium", "LabCholesterol", "LabTriglycerides", "LabHDL", "LabGlucose", "DailyAlcoholicDrinks", "Diabetic","Prediabetic", "WeeklyMealsNotHomePrepared", "HighBP", "Overweight", "CongestiveHeartFailure", "CoronaryHeartDisease", "Stroke", "SedentaryMins", "SmokedCigarettes")

```

### Drop diabetic variable because only 1 value after filtering
```{r}
bp = select(bp,-c(Diabetic))
```
### Splitin dataset: 75% training, 25 testing
```{r}
set.seed(6573)
select.bp <- sample(1:nrow(bp), 0.75*nrow(bp))
bp <- bp[select.bp,]
bp_test <- bp[-select.bp,]
```



### Breaking data down into two sets: one for diastolic and one systolic
```{r}
bp_d = subset(bp, select= -c(BPAvgSys))
bp_s = subset(bp, select= -c(BPAvgDias))
```

### Creating histograms for average diastolic and average systolic blood pressure
```{r}
hist(bp$BPAvgSys, breaks =40)
hist(bp$BPAvgDias, breaks=40)
```

### Loop to create scatterplots of average systolic blood pressure against all other numerical variables
```{r}
numerical = subset(bp, select = c(Age, Calcium, Magnesium, Sodium, Potassium, Alcohol, Weight, BMI, Waist, LabCalcium, LabCholesterol, LabTriglycerides, LabHDL, LabGlucose, DailyAlcoholicDrinks, WeeklyMealsNotHomePrepared, SedentaryMins))
i = 1
while (i <= length(names(numerical))) {
  plot(bp$BPAvgSys, numerical[,i], xlab = "BP Avg Systolic", ylab = names(numerical)[i])
  i = i + 1
}
```

### Loop to create scatterplots of average diastolic blood pressure against all other numerical variables
```{r}

i=1
while (i <= length(names(numerical))) {
  plot(bp$BPAvgDias, numerical[,i], xlab = "BP Avg Diastolic", ylab = names(numerical)[i])
  i = i + 1
}

```

### Loop to create boxlots of average systolic blood pressure against all categorical variables.
```{r}
characs = subset(bp, select = -c(Age, Calcium, Magnesium, Sodium, Potassium, Alcohol, Weight, BMI, Waist, LabCalcium, LabCholesterol, LabTriglycerides, LabHDL, LabGlucose, DailyAlcoholicDrinks, WeeklyMealsNotHomePrepared, SedentaryMins, BPAvgSys, BPAvgDias))
i = 1
while (i <= length(names(characs))) {
  boxplot(bp$BPAvgSys ~ characs[,i], xlab = "BP Avg Systolic", ylab = names(characs)[i])
  i = i + 1
}
```

### Loop to create boxlots of average diastolic blood pressure against all categorical variables
```{r}
i = 1
while (i <= length(names(characs))) {
  boxplot(bp$BPAvgDias ~ characs[,i], xlab = "BP Avg Diastolic", ylab = names(characs)[i])
  i = i + 1
}
```

### Correlation coefficients for average systolic blood pressure
```{r}
cor(bp, method = "pearson")[,"BPAvgSys"]
```

### Correlation coefficients for average diastolic blood pressure
```{r}
cor(bp_d, method = "pearson")[,"BPAvgDias"]
```

### First full linear model for average systolic blood pressure
```{r}
full_sys = lm(BPAvgSys ~ as.factor(Sex)+Age+as.factor(Race)+as.factor(Pregnant)+as.factor(InfantsInHousehold)+as.factor(ChildrenInHousehold)+as.factor(SeniorsInHousehold)+as.factor(Income)+as.factor(SaltUsed)+as.factor(SpecialDiet)+Calcium+Magnesium+Sodium+ Potassium + Alcohol + BMI + Waist + LabCalcium + LabCholesterol + LabTriglycerides + LabHDL + LabGlucose + DailyAlcoholicDrinks + as.factor(Prediabetic) + WeeklyMealsNotHomePrepared + as.factor(HighBP) + as.factor(Overweight) + as.factor(CongestiveHeartFailure)+as.factor(CoronaryHeartDisease)+as.factor(Stroke)+SedentaryMins + as.factor(SmokedCigarettes), data=bp_s)
summary(full_sys)
```

### VIF computed to determine if there is any multi-collinearity in systolic model.  Weight removed because VIF > 10 due to multicollinearity with Waist and BMI.
```{r}
library(car)
```
```{r}
vif(full_sys)
```


### Computation of influential points and studentized vs hat value plot for systolic model. 27 outliers were removed.
```{r influencial, eval=TRUE}
# compute influential points statistics
influence.measures(full_sys)
# print out only observations that may be influential 
summary(influence.measures(full_sys))

plot(rstudent(full_sys)~hatvalues(full_sys))
abline(h =-3, col="red")  # add cutoff line
abline(h = 3, col="red")  # add cutoff line

LargeResiduals <- abs(rstudent(full_sys)) > 3
bp_s_less_out <- bp_s[!LargeResiduals, ]
nrow(bp_s)-nrow(bp_s_less_out)
```

### New systolic model with influential points removed
```{r influencial removed model, eval=TRUE}
full_sys_new = lm(BPAvgSys ~ as.factor(Sex)+Age+as.factor(Race)+as.factor(Pregnant)+as.factor(InfantsInHousehold)+as.factor(ChildrenInHousehold)+as.factor(SeniorsInHousehold)+as.factor(Income)+as.factor(SaltUsed)+as.factor(SpecialDiet)+Calcium+Magnesium+Sodium+ Potassium + Alcohol + BMI + Waist + LabCalcium + LabCholesterol + LabTriglycerides + LabHDL + LabGlucose + DailyAlcoholicDrinks + as.factor(Prediabetic) + WeeklyMealsNotHomePrepared + as.factor(HighBP) + as.factor(Overweight) + as.factor(CongestiveHeartFailure)+as.factor(CoronaryHeartDisease)+as.factor(Stroke)+SedentaryMins + as.factor(SmokedCigarettes), data=bp_s_less_out)

plot(rstudent(full_sys_new)~hatvalues(full_sys_new))
abline(h =-3, col="red")  # add cutoff line
abline(h = 3, col="red")  # add cutoff line
```

### Summary of new systolic model
```{r summary full_sys_new, eval=TRUE}
summary(full_sys_new)
```
```{r stepwise influencial removed , eval=TRUE}
step(full_sys_new, direction = "both", trace=FALSE)
```
### Backward stepwise regression reran on new systolic model
```{r stepwise influencial removed 1 , eval=TRUE}
step(full_sys_new, direction = "backward", trace=FALSE)
```

### Forward stepwise regression
```{r}
Base = lm(BPAvgSys ~ 1, data=bp_s)
step(Base, scope=formula(full_sys),direction = "forward", trace=FALSE)
```


### Stepwise model for new systolic model based on results both from forward and backward stepwise regression
```{r stepwise influencial removed model, eval=TRUE}
sys_less_out = lm(formula = BPAvgSys ~ as.factor(Sex) + Age + as.factor(Race) + 
    as.factor(SeniorsInHousehold) + as.factor(Income) + as.factor(SaltUsed) + 
    as.factor(SpecialDiet) + Calcium + Sodium + Alcohol + BMI + 
    LabCalcium + LabCholesterol + LabTriglycerides + LabHDL + 
    LabGlucose + as.factor(CongestiveHeartFailure) + as.factor(CoronaryHeartDisease), 
    data = bp_s_less_out)
summary(sys_less_out)
```

### Residual and Q-Q Plots for new systolic model
```{r residual-assumptions, eval=TRUE}
#residuals vs fitted values plot
plot(fitted(full_sys), rstandard(full_sys))
abline(a=0,b=0,col='red')

#residuals vs independent variables
plot(bp_s$Age,  rstandard(full_sys), main="Age vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s$Income,  rstandard(full_sys), main="Income vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s$BMI,  rstandard(full_sys), main="Wealth vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s$LabTriglycerides,  rstandard(full_sys), main="Triglycerides vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s$Alcohol,  rstandard(full_sys), main="Alcohol vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s$WeeklyMealsNotHomePrepared,  rstandard(full_sys), main="WeeklyMealsNotHomePrepared vs residuals plot for M2")
abline(a=0, b=0,col='red')

#normal probability plot of residuals
qqnorm(rstandard(full_sys))
qqline(rstandard(full_sys),col=2)
```

### Based on above residual plots, three variables did not appear to have randoly scattered patterns. We attempted to transform these variables.
```{r squared model, eval=TRUE}

Alcohol2 = bp_s$Alcohol^2
LabTriglycerides2 = bp_s$LabTriglycerides^2
WeeklyMealsNotHomePrepared2 = bp_s$WeeklyMealsNotHomePrepared^2

# attach it to bp_s dataset
bp_s=cbind(bp_s, Alcohol2,LabTriglycerides2, WeeklyMealsNotHomePrepared2)

full_sys_transformed = lm(BPAvgSys ~ as.factor(Sex)+Age+as.factor(Race)+as.factor(Pregnant)+as.factor(InfantsInHousehold)+as.factor(ChildrenInHousehold)+as.factor(SeniorsInHousehold)+as.factor(Income)+as.factor(SaltUsed)+as.factor(SpecialDiet)+Calcium+Magnesium+Sodium+ Potassium + Alcohol + Alcohol2 + BMI + Waist + LabCalcium + LabCholesterol + LabTriglycerides+ LabTriglycerides2 + LabHDL + LabGlucose + DailyAlcoholicDrinks + as.factor(Prediabetic) + WeeklyMealsNotHomePrepared + WeeklyMealsNotHomePrepared2 + as.factor(HighBP) + as.factor(Overweight) + as.factor(CongestiveHeartFailure)+as.factor(CoronaryHeartDisease)+as.factor(Stroke)+SedentaryMins + as.factor(SmokedCigarettes), data=bp_s)
summary(full_sys_transformed)
```

#### The adjusted R-squared improves decresed//from 0.3076 after transformation to 0.3103

### Residual and Q-Q Plots recreated after transformations

```{r residual-assumptions after transform, eval=TRUE}
#residuals vs fitted values plot
plot(fitted(full_sys_transformed), rstandard(full_sys_transformed))
abline(a=0,b=0,col='red')

plot(bp_s$LabTriglycerides2,  rstandard(full_sys_transformed), main="Triglycerides vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s$Alcohol2,  rstandard(full_sys_transformed), main="Alcohol vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s$WeeklyMealsNotHomePrepared2,  rstandard(full_sys_transformed), main="WeeklyMealsNotHomePrepared vs residuals plot for M2")
abline(a=0, b=0,col='red')

#normal probability plot of residuals
qqnorm(rstandard(full_sys_transformed))
qqline(rstandard(full_sys_transformed),col=2)
```

### Stepwise regression redone after transformation improved model created. This model still contains the outliers identified earlier in the analysis 
```{r stepwise transformed , eval=TRUE}
step(full_sys_transformed, direction = "backward", trace=FALSE)
```

### Stepwise regression model from above output
```{r stepwise squared model, eval=TRUE}
sys_trans = lm(formula = BPAvgSys ~ as.factor(Sex) + Age + as.factor(Race) + 
    as.factor(SeniorsInHousehold) + as.factor(Income) + as.factor(SaltUsed) + 
    as.factor(SpecialDiet) + Magnesium + Alcohol + Alcohol2 + 
    BMI + LabCholesterol + LabTriglycerides + LabTriglycerides2 + 
    LabHDL + as.factor(CongestiveHeartFailure) + as.factor(CoronaryHeartDisease), 
    data = bp_s)
summary(sys_trans)
```

### Full system model created again, but this time without the outliers identified earlier
```{r transformed and less residual, eval=TRUE}
Alcohol2 = bp_s_less_out$Alcohol^2
LabTriglycerides2 = bp_s_less_out$LabTriglycerides^2
WeeklyMealsNotHomePrepared2 = bp_s_less_out$WeeklyMealsNotHomePrepared^2

# attach it to bp_s dataset
bp_s_less_out=cbind(bp_s_less_out, Alcohol2,LabTriglycerides2, WeeklyMealsNotHomePrepared2)

full_sys_tran_less_out = lm(BPAvgSys ~ as.factor(Sex)+Age+as.factor(Race)+as.factor(Pregnant)+as.factor(InfantsInHousehold)+as.factor(ChildrenInHousehold)+as.factor(SeniorsInHousehold)+as.factor(Income)+as.factor(SaltUsed)+as.factor(SpecialDiet)+Calcium+Magnesium+Sodium+ Potassium + Alcohol + Alcohol2 + BMI + Waist + LabCalcium + LabCholesterol + LabTriglycerides+ LabTriglycerides2 + LabHDL + LabGlucose + DailyAlcoholicDrinks + as.factor(Prediabetic) + WeeklyMealsNotHomePrepared + WeeklyMealsNotHomePrepared2 + as.factor(HighBP) + as.factor(Overweight) + as.factor(CongestiveHeartFailure)+as.factor(CoronaryHeartDisease)+as.factor(Stroke)+SedentaryMins + as.factor(SmokedCigarettes), data=bp_s_less_out)
summary(full_sys_tran_less_out)
```


### Stepwise regression on full system transformation model without outliers
```{r stepwise transformed and oulier removed, eval=TRUE}
step(full_sys_tran_less_out, direction = "backward", trace=FALSE)
```

```{r}
Base_less_out = lm(BPAvgSys ~ 1, data=bp_s_less_out)
step(Base_less_out, scope=formula(full_sys_tran_less_out),direction = "forward", trace=FALSE)
```

```{r stepwise transformed and oulier removed model, eval=TRUE}
sys_trans_out_comb = lm(formula = BPAvgSys ~ as.factor(Sex) + Age + as.factor(Race) + 
    as.factor(SeniorsInHousehold) + as.factor(Income) + as.factor(SaltUsed) + 
    Calcium + Sodium + Alcohol + Alcohol2 + BMI + LabCalcium + 
    LabTriglycerides + LabTriglycerides2 + LabHDL + LabGlucose + 
    as.factor(CongestiveHeartFailure) + as.factor(CoronaryHeartDisease), 
    data = bp_s_less_out)
summary(sys_trans_out_comb)
```

#### The adjusted R-squared is highest after transformation and influential large residual removal compared to transformation model with outliers included.

### Create dummy variables for significant variable factors
```{r}
d_Race_4=(bp_s_less_out$Race == 4)*1
d_Race_5=(bp_s_less_out$Race == 5)*1
d_SeniorsinHousehold_1=(bp_s_less_out$SeniorsInHousehold == 1)*1
d_SeniorsinHousehold_2=(bp_s_less_out$SeniorsInHousehold == 2)*1
d_Sex_2=(bp_s_less_out$Sex == 2)*1
d_SaltUsed_2=(bp_s_less_out$SaltUsed == 2)*1
d_SaltUsed_3=(bp_s_less_out$SaltUsed == 3)*1
d_SaltUsed_4=(bp_s_less_out$SaltUsed == 4)*1
d_CongHeartFail_2=(bp_s_less_out$CongestiveHeartFailure == 2)*1
d_CorHeartDisease_2=(bp_s_less_out$CoronaryHeartDisease == 2)*1
```

### Remove insignificant variables
```{r}
sys_trans_out_comb2 = lm(formula = BPAvgSys ~ as.factor(Sex) + Age + d_Race_4 + d_Race_5 + 
    d_SeniorsinHousehold_1 + d_SeniorsinHousehold_2 + as.factor(SaltUsed) + 
    Alcohol + Alcohol2 + BMI +
    LabTriglycerides + LabTriglycerides2 + LabHDL + 
    as.factor(CongestiveHeartFailure) + as.factor(CoronaryHeartDisease), 
    data = bp_s_less_out)
summary(sys_trans_out_comb2)
```

### Residuals and Q-Q plots reran on the latest model

```{r residual-assumptions after transform and large residual removal, eval=TRUE}
#residuals vs fitted values plot
plot(fitted(sys_trans_out_comb2), rstandard(sys_trans_out_comb2))
abline(a=0,b=0,col='red')

plot(bp_s_less_out$LabTriglycerides2,  rstandard(sys_trans_out_comb2), main="Triglycerides vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s_less_out$BMI,  rstandard(sys_trans_out_comb2), main="BMI vs residuals plot for M2")
abline(a=0, b=0,col='red')
plot(bp_s_less_out$Alcohol2,  rstandard(sys_trans_out_comb2), main="Alcohol vs residuals plot for M2")
abline(a=0, b=0,col='red')

#normal probability plot of residuals
qqnorm(rstandard(sys_trans_out_comb2))
qqline(rstandard(sys_trans_out_comb2),col=2)
```

### Check against interactions with Sex since boxplot was the only one that showed noticeable differences in medians and IQRs. Perform backward stepwise regression for possible model.
```{r}
sys_trans_out_int_full = lm(formula = BPAvgSys ~ as.factor(Sex)*( Age + as.factor(Race) + 
    as.factor(SeniorsInHousehold) + as.factor(Income) + as.factor(SaltUsed) + 
    Calcium + Sodium + Alcohol + Alcohol2 + BMI + LabCalcium + 
    LabTriglycerides + LabTriglycerides2 + LabHDL + LabGlucose + 
    as.factor(CongestiveHeartFailure) + as.factor(CoronaryHeartDisease)), 
    data = bp_s_less_out)
step(sys_trans_out_int_full, direction="backward", trace=FALSE)
```
###   Not much improvement in model when check for possible interactions and remove insignificant values.
```{r}
sys_trans_out_int1 = lm(formula = BPAvgSys ~ Age + d_Race_4 + d_Race_5 + 
    d_SeniorsinHousehold_1 + d_SeniorsinHousehold_2 + as.factor(SaltUsed) + 
    Alcohol + Alcohol2 + BMI + 
    LabTriglycerides + LabTriglycerides2 + LabHDL + 
    as.factor(CongestiveHeartFailure) + as.factor(CoronaryHeartDisease) + 
    as.factor(Sex):LabHDL, data = bp_s_less_out)
summary(sys_trans_out_int1)
```

### Cross-validation for full_sys_tran_less_out and sys_trans_out_int1 to choose simpler model.
```{r}
Alcohol2 = bp_test$Alcohol^2
LabTriglycerides2 = bp_test$LabTriglycerides^2
WeeklyMealsNotHomePrepared2 = bp_test$WeeklyMealsNotHomePrepared^2

d_Race_4=(bp_test$Race == 4)*1
d_Race_5=(bp_test$Race == 5)*1
d_SeniorsinHousehold_1=(bp_test$SeniorsInHousehold == 1)*1
d_SeniorsinHousehold_2=(bp_test$SeniorsInHousehold == 2)*1

bp_test=cbind(bp_test, Alcohol2,LabTriglycerides2,WeeklyMealsNotHomePrepared2,d_Race_4,d_Race_5,d_SeniorsinHousehold_1,d_SeniorsinHousehold_2)
y_pred <- predict.glm(full_sys_tran_less_out, bp_test)
y_pred2 <- predict.glm(sys_trans_out_int1, bp_test)
y_obs<-bp_test[,"BPAvgSys"]
```
### Cross-validation statistics
```{r}
# Compute RMSE of prediction errors 
rmse_m1 <- sqrt((y_obs - y_pred)%*%(y_obs - y_pred)/nrow(bp_test)) 
rmse_m1
# Compute mean absolute error 
mae_m1<-mean(abs(y_obs - y_pred)) 
mae_m1
# Compute mean percentage absolute error 
mape_m1<-mean(abs((y_obs - y_pred)/y_obs))*100 
mape_m1
```

```{r}
# Compute RMSE of prediction errors 
rmse_m2 <- sqrt((y_obs - y_pred2)%*%(y_obs - y_pred2)/nrow(bp_test)) 
rmse_m2
# Compute mean absolute error 
mae_m2<-mean(abs(y_obs - y_pred2)) 
mae_m2
# Compute mean percentage absolute error 
mape_m2<-mean(abs((y_obs - y_pred2)/y_obs))*100 
mape_m2
```



```{r}
# compute cross-validated R^2_pred 
r2_pred = cor(cbind(y_obs,y_pred))**2 
r2_train = summary(full_sys_tran_less_out)$r.squared 
diffr2_m1=abs(r2_train-r2_pred) 
#print difference of cross-validate R2 and R2 
diffr2_m1[1,2]

```

```{r}
# compute cross-validated R^2_pred 
r2_pred2 = cor(cbind(y_obs,y_pred2))**2 
r2_train2 = summary(sys_trans_out_int1)$r.squared 
diffr2_m2=abs(r2_train2-r2_pred2) 
#print difference of cross-validate R2 and R2 
diffr2_m2[1,2]

```
